Determining the existance and direction of causality in the face of measurement error

### 2016-05-06 docs/manuscript.Rmd
updating manuscript

With rapid acceleration in technological advances that enable us to measure the human phenome, we are presented with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.

An established tool that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the association by randomly splitting the sample into two groups, applying the exposure to one group and leaving the other as a control. The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship. In practical terms randomisation is often impossible to implement in such a controlled manner, but the principles can be employed using existing cross sectional data through the use of genetic instruments.

@@

@@

 Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods, which have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

>>>>

Rapid acceleration in technological advances have occurred over the last decade that enable us to measure the human phenome in increasing detail. This presents us with the statistical challenge of correctly defining the causal architectures that underlie these high dimensional data. For example, constructing gene networks, identifying metabolic perturbations that lead to disease, or understanding the influence of epigenomic influences on gene all depend on understanding the causal nature of any putative observational associations. Despite the ubiquity of the question, cross-sectional study designs are notoriously susceptible to pitfalls in making causal inference, with simple regression based techniques unable to distinguish a true causal association from reverse causation or confounding.

An established tool that can resolve this problem is randomisation. Given the hypothesis that an exposure is causally related to an outcome, randomisation can be employed to assess the association by randomly splitting the sample into two groups, applying the exposure to one group and leaving the other as a control. The association between the exposure and the outcome in this setting is then guaranteed to be a robust estimate of the causal relationship. This is the theoretical basis behind randomised control trials, but in practical terms randomisation is often impossible to implement in such a controlled manner. The principle, however, can be employed using existing cross sectional data through the use of genetic variants (instruments), which have the potentially useful property of proxying for random allocation of exposure levels by virtue of whether a causal allele was inherited. There have been two broad statistical approaches to exploiting the properties of genetic instruments: mediation and Mendelian randomisation. Comparisons are often made between the two, but a crucial difference in how they behave is in their robustness in the presence of measurement error. Here we will introduce the two basic approaches and summarise why measurement error is a serious issue that needs to be considered in method choice for causal analysis.

@@

This simplistic scenario, however, is seldom borne out in reality. Alternative models of reverse causation or confounding are often indistinguishable from the hypothesised model of the independent variable causing the dependent variable via the mediator. A proposed solution to this is to use a genetic instrument as the independant variable. Supposing an exposure has a causal effect on an outcome, then an ‘instrument’ that causes the exposure (a single nucleotide polymorophism (SNP), for example) should also influence the outcome. Therefore the influence of the instrument on the outcome conditional on the exposure should be zero. This forms the basis of a number of methods, such as the regression-based causal inference test (CIT) [@Millstein2009], an SEM implementation in the NEO software [@Aten2008], and various Bayesian methods. They have been employed by a number of recent publications that make causal inferences, often in large scale ‘omics datasets [@Koestler2014; @Liu2013; @Waszak2015].

@@
